MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-04-12 latex/paper/paper.tex
lutffpair

    In this paper we will be working with the Xilinx 7-Series devices which follows the heterogenous island-style architecture. 
    The 7-Series architecture first released in 2010 and was superceded by the Ultrascale architecture which made several improvements over the 7-Series. 
    Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility to beginner users and compatibility with open source tooling.

    % Figure~\ref{fig:hierarchy} illustrates the architectural hierarchy of an Ultrascale Xilinx FPGA. 
    % The architecture hierarchy is the same for the 7-Series devices. 
    Figure \ref{fig:hierarchy_4} illustrates the architectural hierarchy of a 7-Series Xilinx FPGA. 
    At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs). 
    These BELs, which include LUTs, FFs, BRAMs, DSPs, and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.

@@

    Note that in the Xilinx architectures, "CLBs" refer to CLB Tiles, each containing two "SLICE" Sites, with each SLICE Site containing 8 LUTs, 8 Flip Flop, and 1 CARRY BEL.
    These SLICEs facilitate the core reprogrammability of the FPGA.

@@

\section{Placement Implementation}

    With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
    Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement.

@@

    \subsection{PrePacking}

        In the prepacking stage, we traverse the raw post-synthesis EDIF Netlist to identify common recurring Cell patterns. 
        We do this because there are certain desirable cell configurations that should be clustered together to exploit some proximity characteristics of the FPGA architecture. 
        For example, in Figure \ref{fig:intrasite_intersite}
        Doing so will innately improve wirelength minimization. 
        There are also some cell configurations that must necessarily be placed in certain ways in relation to each other to ensure legality within the architecture constraints.

        \begin{figure}
            \centering
            \includegraphics[width=8.0cm]{figures/intersite_intrasite.png}
            \caption{
                \textbf{Upper Left}: A LUT-FF cell pair in the netlist.
                \textbf{Upper Right}: The LUT-FF pair placed within the same Site.
                \textbf{Bottom}: The LUT-FF pair placed across different Sites.
            }
            \label{fig:intrasite_intersite}
        \end{figure}

        For example, in many HDL designs, LUTs and FFs often come in pairs, as is the case in Figure \ref{fig:intrasite_intersite}

@@

        \subsection{LUT-FF Pairs}
            Identify unique CE-SR net pairs. 
            All FF cells in a Site must share the same CE and SR nets.

>>>>

    In this paper we will be working with the Xilinx 7-Series devices which follows the heterogenous island-style architecture. 
    The 7-Series architecture first released in 2010 and was superceded in 2013 by the Ultrascale architecture which made several improvements over the 7-Series. 
    Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility to beginner users and compatibility with open source tooling.

    Figure \ref{fig:hierarchy_4} illustrates the architectural hierarchy of a 7-Series Xilinx FPGA. 
    At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs). 
    These BELs, which include LUTs, FFs, BRAMs, DSPs, and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.

@@

    Note that in the Xilinx architectures, "CLBs" refer to CLB Tiles, each containing two "SLICE" Sites, with each SLICE Site containing 8 LUTs, 8 Flip Flop, and 1 CARRY BEL. 
    These SLICE Sites facilitate the core programmable logic of the FPGA.

@@

\section{Placement Implementation}
    With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
    Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement.

@@

    \subsection{PrePacking}
        In the prepacking stage, we traverse the raw post-synthesis EDIF Netlist to identify common recurring Cell patterns. 
        We do this because there are certain desirable cell configurations that should be clustered together to exploit the physical organization of the BELs and Sites of the architecture. 
        Doing so will innately improve wirelength minimization. 
        There are also some cell configurations that must necessarily be placed in certain ways in relation to each other to ensure legality within the architecture constraints.

        \subsection{LUT-FF Pairs}
            In the vast majority of HDL designs, LUTs and FFs will be synthesized in pairs, as is the case in Figure \ref{fig:lut_ff_pair}.
            Shown in Figure \ref{fig:intrasite_intersite} are two possible placements for this cell net. 
            On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
            On the bottom left, the cells are placed within the same Site, taking advantage of the intra-site routing and avoiding costly inter-site routing.

            Identify unique CE-SR net pairs. 
            All FF cells in a Site must share the same CE and SR nets.

        \begin{figure}[H]
            \centering
            \includegraphics[width=8.0cm]{figures/lut_ff_pair.png}
            \caption{
                Example of a LUT-FF Cell Pair as shown the Vivado netlist viewer. 
                FDRE is a type of Flip Flop. 
            }
            \label{fig:lut_ff_pair}
        \end{figure}

        \begin{figure}[H]
            \centering
            \includegraphics[width=8.0cm]{figures/intersite_intrasite.png}
            \caption{
                \textbf{Upper Left}: A LUT-FF cell pair in the netlist.
                \textbf{Upper Right}: The LUT-FF pair placed within the same Site.
                \textbf{Bottom}: The LUT-FF pair placed across different Sites.
            }
            \label{fig:intrasite_intersite}
        \end{figure}

@@
